Millions of Type Errors

### 2023-05-03 oopsla23/paper.tex
mode vs. change in error, macros for the modes

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}

@@

To opt in to type error reporting, creators set a \emph{mode}
for each script, which is one of:
\begin{itemize}
\item \emph{nocheck}: only syntax errors are reported,
\item \emph{nonstrict}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
\item \emph{strict}: all syntax errors and type errors, are reported.
\end{itemize}
As an example of nonstrict mode, the following program only reports one error:
\begin{verbatim}
--!nonstrict
local x = { p = 5, q = nil }
if condition then x.q = 7 end
local y = x.p + x.q --> no type error
local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
but in strict mode it reports two:
\begin{verbatim}
--!strict
local x = { p = 5, q = nil }
if condition then x.q = 7 end
local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
In cases like this, where it is undecidable whether there will be a run-time error,
strict mode errs on the side of reporting an error, and nonstrict mode errs on
the side of suppressing the error.

@@

Both modes are opt-in. Creators have the option to make nonstrict mode
the default rather than nocheck mode.

Even in nocheck mode, \anon{Roblox Studio} performs type inference, since
the results are needed by type-directed tooling such as autocomplete and
API documentation. This behind-the-scenes typechecking is always performed
in strict mode, since it is important that the inferred types be as precise
as possible. The type errors produced by this pass are always discarded,
so the verbosity of strict mode is not an issue.

@@

\begin{description}
  \item[D1] metadata: uid, timestamp, session duration
  \item[D2] mode (strict / nonstrict / nocheck)
  \item[D3] a 1-bit reason for sending: did the user switch module?
    \subitem we get logs for two reasons: (1) switch module, (2) randomly with
    low probability after typechecking
  \item[D4] size of codebase: num files, num lines
  \item[D5] size of edit region
  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in region
  \item[D7] specific errors in the edit region: in the edit region, for old and
    curr states, how many type errors for each type error code? (There is 1
    edit region per log.)
    \subitem Examples, if the old state had one type-A error:
    \begin{itemize}
      \item If the latest edits touched it and fixed it, the next log says 1
        old type-A and 0 new type-A
      \item touched it, didn't fix => 1 old-A, 1 new-A
      \item didn't touch (fixed or not) => 0 old-A, 0 new-A
    \end{itemize}
  \item[D8] ForcedStrict "what if" errors for nocheck: specific errors for the
    edit region computed as if the user had strict mode enabled (even though they
    have no-check enabled)
    \subitem This info is available because it gives better hints for
    autocomplete, because it assumes precise types for ALL libraries (rather
    than Any)
  \item[D9] TooComplexErrors: project-wide count for this one specific error type
\end{description}

@@

\begin{itemize}
  \item How often does the typechecker give up with CodeTooComplex errors?
  \subitem [[ For context: the type system is designed to give useful feedback
    to all sorts of code, whether strict-typed, nonstrict-typed, or nocheck.
    But it has an escape hatch: CTC. Hope it's rarely needed. ]]
    \begin{itemize}
      \item Count the overall CTC errors (in each mode) for a coarse view of
        how rare they are. [D9 of course, D8 to compare nocheck code with
        strict-mode code via the same typechecker]
      \subitem Note: We can't look at the code to learn why the errors
        happened. Can only look for trends and tell the Roblox team where to
        look.
      %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around. [D9 for the count, D8 for nocheck vs strict, D7 because CTC might show up in the edit region]
    \end{itemize}
  \item How helpful are highlighted regions?
    \begin{itemize}
      \item How often do edits in a highlighted region remove type errors? How
        often do they add new ones?
      \subitem [D7 compare specific old vs specific new]
      \subitem [may want to filter using D2 + D5]
      \item How often do edits outside the region remove type errors? How often do they add new ones?
      \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
      \item Which specific errors seem easy / hard to fix via highlights?
      \subitem [D6 removed by edit => easy; D6 survive edit => hard]
    \end{itemize}
  \item In what ways do types improve quality of life?
  \begin{itemize}
    \item Does nocheck code have many more latent errors than the others?
    \subitem [D8 latent errors come from the "what if" analysis]
  \end{itemize}
\end{itemize}

% What to study?
% 
% Sanity checks
% How large are edit regions? Plot the distribution.
% How long between logs? Long times & very short times are problematic.
% General Roblox trends
% How many sessions use strict / nonstrict / nocheck?
% Expect few strict/nonstrict, lots of nocheck
% How often does mode change within a session?
% Expect few of these b/c error reports are easy to ignore; non-intrusive
% If a down-switch happens, expect the current #errors to be high
% If an up-switch happens, hope to stay up
% How many CodeTooComplex errors arise, overall?
% Expect very few. Hopefully zero.
% How often do module switches happen?
% Expect not-too-many, with some logs in between each time
% Compare strict vs. nonstrict vs. nocheck
% Get median # errors across all snapshots.
% Expect nocheck median to be higher than the others.
% Plot #errors vs time, or vs codebase size
% Expect nocheck errors to keep going up. For other modes, #errors should stay low.
% General type error trends
% Plot the distribution of #errors across all edit regions
% Expect different counts based on error-kind. SEE BELOW
% Count specific errors that get fixed. For each kind of error, how many times is it old-in-region but not new-in-region? Compare to #old-in-region totals.
% Expect all errors to have a decent fix rate.
% If one error has a lower % than others, its error region may be misleading. Or, maybe it's hard to fix
% Type errors vs. edit region
% Compare #old vs. #new in region
% Expect a decrease, #old > #new
% If there's an increase, look for an explanation. Could be a huge edit region. Could be a 1-line region that we captured 2x in a row. Could be that this user is simply ignoring type errors.
% Compare [#old-in-region vs. (total #old - total #new)]
% If #old-in-region = 0, then hope the total # increases
% Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
% If #old-in-region > 0, then hope the total # decreases
% Increase (in region or otherwise) => made things worse

@@

\begin{figure}[t]
  \begin{tabular}{rl}
   5,001 & total logs \\
         & 4,471 nocheck + 506 nonstrict + 24 strict \\
         & 1,628 from module switches \\
   182,247 & total forced strict type errors \\
         & {83,005 in module, 5,113 in edit regions} \\
    1,521 & total type errors \\
         & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
    \\[2ex]

    1,146 & sessions \\
    & \begin{tabular}{l}
        1,143 single mode = 1046 nocheck + 94 nonstrict + 3 strict \\
        detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
      \end{tabular}
  \end{tabular}

@@

\begin{figure}[t]
  %% TODO stacked bars?
  %% TODO align old;curr
  \begin{tabular}{lrrrrr}
    Mode             & \# Rows & E         & Any Error & In Module & In Region \\
      & & & & (old ; curr) &  (old ; curr) \\\midrule
    \code{nocheck}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
                     &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
    \code{nonstrict} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
                     &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
    \code{strict}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
                     &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
  \end{tabular}

@@

  \begin{tabular}{lrrrr}
    & Local Remove & Local Add & Nonlocal Remove & Nonlocal Add \\
    Mode
    & \#err (count)
    & \#err (count)
    & \#err (count)
    & \#err (count) \\\midrule
    \code{nocheck}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
    \code{nonstrict} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
    \code{strict}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
  \end{tabular}

@@

@@

How many module switches exit a module with errors?
nocheck 24, nonstrict 9, strict 3.
Rare?
More common for forcedstrict:
 nocheck 190, nonstrict 16, strict 2.

How many module switches go to a module with errors?
nocheck 70, nonstrict 9, strict 4
For FS: nocheck 400, nonstrict 39, strict 6.

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}

@@

To opt in to type error reporting, creators set a \emph{mode}
for each script, which is one of:
\begin{itemize}
  \item \mnocheck{}: only syntax errors are reported,
  \item \mnonstrict{}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
  \item \mstrict{}: all syntax errors and type errors, are reported.
\end{itemize}
As an example of nonstrict mode, the following program only reports one error:
\begin{verbatim}
--!nonstrict
local x = { p = 5, q = nil }
if condition then x.q = 7 end
local y = x.p + x.q --> no type error
local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
but in strict mode it reports two:
\begin{verbatim}
--!strict
local x = { p = 5, q = nil }
if condition then x.q = 7 end
local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
In cases like this, where it is undecidable whether there will be a run-time error,
strict mode errs on the side of reporting an error, and nonstrict mode errs on
the side of suppressing the error.

@@

Both modes are opt-in. Creators have the option to make nonstrict mode
the default rather than \mnocheck{} mode.

Even in \mnocheck{} mode, \anon{Roblox Studio} performs type inference, since
the results are needed by type-directed tooling such as autocomplete and
API documentation. This behind-the-scenes typechecking is always performed
in strict mode, since it is important that the inferred types be as precise
as possible. The type errors produced by this pass are always discarded,
so the verbosity of strict mode is not an issue.

@@

\begin{description}
  \item[D1] metadata: uid, timestamp, session duration
  \item[D2] mode (\mnocheck{}, \mnonstrict{}, \mstrict{})
  \item[D3] a 1-bit reason for sending: did the user switch module?
    \subitem we get logs for two reasons: (1) switch module, (2) randomly with
    low probability after typechecking
  \item[D4] size of codebase: num files, num lines
  \item[D5] size of edit region
  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in region
  \item[D7] specific errors in the edit region: in the edit region, for old and
    curr states, how many type errors for each type error code? (There is 1
    edit region per log.)
    \subitem Examples, if the old state had one type-A error:
    \begin{itemize}
      \item If the latest edits touched it and fixed it, the next log says 1
        old type-A and 0 new type-A
      \item touched it, didn't fix => 1 old-A, 1 new-A
      \item didn't touch (fixed or not) => 0 old-A, 0 new-A
    \end{itemize}
  \item[D8] ForcedStrict "what if" errors for \mnocheck{}: specific errors for the
    edit region computed as if the user had strict mode enabled (even though they
    have no-check enabled)
    \subitem This info is available because it gives better hints for
    autocomplete, because it assumes precise types for ALL libraries (rather
    than Any)
  \item[D9] TooComplexErrors: project-wide count for this one specific error type
\end{description}

@@

\begin{itemize}
  \item How often does the typechecker give up with CodeTooComplex errors?
  \subitem [[ For context: the type system is designed to give useful feedback
    to all sorts of code, whether strict-typed, nonstrict-typed, or \mnocheck{}.
    But it has an escape hatch: CTC. Hope it's rarely needed. ]]
    \begin{itemize}
      \item Count the overall CTC errors (in each mode) for a coarse view of
        how rare they are. [D9 of course, D8 to compare \mnocheck{} code with
        strict-mode code via the same typechecker]
      \subitem Note: We can't look at the code to learn why the errors
        happened. Can only look for trends and tell the Roblox team where to
        look.
        %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around. [D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit region]
    \end{itemize}
  \item How helpful are highlighted regions?
    \begin{itemize}
      \item How often do edits in a highlighted region remove type errors? How
        often do they add new ones?
      \subitem [D7 compare specific old vs specific new]
      \subitem [may want to filter using D2 + D5]
      \item How often do edits outside the region remove type errors? How often do they add new ones?
      \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
      \item Which specific errors seem easy / hard to fix via highlights?
      \subitem [D6 removed by edit => easy; D6 survive edit => hard]
    \end{itemize}
  \item In what ways do types improve quality of life?
  \begin{itemize}
    \item Does \mnocheck{} code have many more latent errors than the others?
    \subitem [D8 latent errors come from the "what if" analysis]
  \end{itemize}
\end{itemize}

% What to study?
% 
% Sanity checks
% How large are edit regions? Plot the distribution.
% How long between logs? Long times & very short times are problematic.
% General Roblox trends
% How many sessions use strict / nonstrict / \mnocheck{}?
% Expect few strict/nonstrict, lots of \mnocheck{}
% How often does mode change within a session?
% Expect few of these b/c error reports are easy to ignore; non-intrusive
% If a down-switch happens, expect the current #errors to be high
% If an up-switch happens, hope to stay up
% How many CodeTooComplex errors arise, overall?
% Expect very few. Hopefully zero.
% How often do module switches happen?
% Expect not-too-many, with some logs in between each time
% Compare strict vs. nonstrict vs. nocheck
% Get median # errors across all snapshots.
% Expect nocheck median to be higher than the others.
% Plot #errors vs time, or vs codebase size
% Expect nocheck errors to keep going up. For other modes, #errors should stay low.
% General type error trends
% Plot the distribution of #errors across all edit regions
% Expect different counts based on error-kind. SEE BELOW
% Count specific errors that get fixed. For each kind of error, how many times is it old-in-region but not new-in-region? Compare to #old-in-region totals.
% Expect all errors to have a decent fix rate.
% If one error has a lower % than others, its error region may be misleading. Or, maybe it's hard to fix
% Type errors vs. edit region
% Compare #old vs. #new in region
% Expect a decrease, #old > #new
% If there's an increase, look for an explanation. Could be a huge edit region. Could be a 1-line region that we captured 2x in a row. Could be that this user is simply ignoring type errors.
% Compare [#old-in-region vs. (total #old - total #new)]
% If #old-in-region = 0, then hope the total # increases
% Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
% If #old-in-region > 0, then hope the total # decreases
% Increase (in region or otherwise) => made things worse

@@

\begin{figure}[t]
  \begin{tabular}{rl}
   5,001 & total logs \\
         & 4,471 \mnocheck{} + 506 \mnonstrict{} + 24 \mstrict{} \\
         & 1,628 from module switches \\
   182,247 & total forced strict type errors \\
         & {83,005 in module, 5,113 in edit regions} \\
    1,521 & total type errors \\
         & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
    \\[2ex]

    1,146 & sessions \\
    & \begin{tabular}{l}
      1,143 single mode = 1046 \mnocheck{} + 94 \mnonstrict{} + 3 \mstrict{} \\
        detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
      \end{tabular}
  \end{tabular}

@@

\begin{figure}[t]
  %% TODO stacked bars?
  %% TODO align old;curr
  \begin{tabular}{lrrrrr}
    Mode             & \# Rows & E         & Any Error & In Module & In Region \\
      & & & & (old ; curr) &  (old ; curr) \\\midrule
    \mnocheck{}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
                  &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
    \mnonstrict{} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
                  &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
    \mstrict{}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
                  &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
  \end{tabular}

@@

  \begin{tabular}{lrrrr}
    & Local Remove & Local Add & Nonlocal Remove & Nonlocal Add \\
    Mode
    & \#err (count)
    & \#err (count)
    & \#err (count)
    & \#err (count) \\\midrule
    \mnocheck{}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
    \mnonstrict{} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
    \mstrict{}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
  \end{tabular}

@@

FILL why study, what implications?

@@

How many module switches exit a module with errors?
\mnocheck{} 24, \mnonstrict{} 9, \mstrict{} 3.
Rare?
More common for forcedstrict:
 \mnocheck{} 190, \mnonstrict{} 16, \mstrict{} 2.

How many module switches go to a module with errors?
\mnocheck{} 70, nonstrict 9, strict 4
For FS: \mnocheck{} 400, \mnonstrict{} 39, \mstrict{} 6.

\subsection{Types and Quality of Life}

FILL error density: number vs region size

\begin{figure}[h]
  \begin{tabular}{llrrr}
    Mode & E & Add & Keep & Remove \\\midrule
    \mnocheck{}   & \code{te} & - & - & - \\
                  & \code{fs} & - & - & - \\
    \mnonstrict{} & \code{te} & 16 & 18 & 21 \\
                  & \code{fs} & 19 & 18 & 18 \\
    \mstrict{}    & \code{te} & - & - & - \\
                  & \code{fs} & - & - & - \\
  \end{tabular}

  \caption{Does mode influence the change in total error count?}
  \label{f:edit-deltas}
\end{figure}

TODO scribble figures ...
